---
title: "1. Data Visualization"
author: "Matheus C. Negrellos"
date: "2025-07-15"
output: html_document
---

```{r Load package(s), message=FALSE, warning=FALSE}
library(tidyverse)
library(palmerpenguins) # Contains body measurements for penguins in 3 islands
library(ggthemes)
```

# 1. Creating a plot

- Central question: “What does the relationship between flipper length and body mass look like in the palmerpenguins dataset?”

- First, we just state the dataset to use in the graph
```{r}
ggplot(data = penguins)
```

- Next, we need to indicate how the information will be visually represented
```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
)
```

- Understanding aesthetic mappings  
  * `mapping` defines how variables in your dataset are linked to plot aesthetics (aes)
  * In the `aes` function we identify which variables to map in each axis

- However, we have not yet defined how to represent the observations from the data on the plot
  * To do so, we need to define a `geom_`
    * It is the geometrical object that a plot uses to represent data
  * Different plots use different types of `geom_`:
    * `geom_bar()`: Bar charts;
    * `geom_line()`: Line charts;
    * `geom_boxplot()`: Boxplots;
    * `geom_point()`: Scatter plots; etc.

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point()
```


# 2. Adding layeers to the plot

- Let's add the variable  `species` to bring more information to our graph
  * We will do it representing species with different colored points (`color = species`)

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species)
) +
  geom_point()
```

- Now, let's add a curve displaying the relationship between body mass and flipper length
  * This is a new geometric object, so we will need a new `geom_`: `geom_smooth()`
    * We also specify that we want the line of best fit based on a linear model: `method = "lm"`
  
```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species)
) +
  geom_point() +
  geom_smooth(method = "lm")
```

- We successfully added the lines, but this is not the final plot we want
  * We want a plot that have just one line for the entire dataset (not one for each species)
  
- What is happenin right now
  * Aesthetic mappings are defined at the **global level** in the `ggplot()`
  * So, it passes down to each of the subsequent `geom` layers of the plot

- However, each `geom_` function can take a its own `mapping` argument
  * If we specify it, this mapping will only apply to that specific `geom_` function
  
- We can see below 2 codes that return the same graph:

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species)
) +
  geom_point()

ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point(mapping = aes(color = species))
```

- However, in the 2nd one, the  `color = species` aesthetic is set only for the `geom_point()` layer
  * So, only the points will receive this color differentiation by species
  * If we add other layers (e.g., `geom_smooth()`), they will not inherit this color mapping
    
```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point(mapping = aes(color = species)) +
  geom_smooth(method = "lm")
```
  
- It is usually a bad idea to represent information just using colors on a plot
  * So, we can also map species to the "shape" aesthetic

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point(mapping = aes(color = species, shape = species)) +
  geom_smooth(method = "lm")
```

- To finally achieve our goal, we just have to enhance the labels of the graph
  * We can use the "labs()" function to create a new layer

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point(aes(color = species, shape = species)) +
  geom_smooth(method = "lm") +
  labs(
    title = "Body mass and flipper length",
    subtitle = "Dimensions for Adelie, Chinstrap, and Gentoo Penguins",
    x = "Flipper length (mm)", y = "Body mass (g)",
    color = "Species", shape = "Species"
  ) +
  scale_color_colorblind() # improves the color palette to be colorblind safe
```

# 3. Exercises (1)

1) How many rows are in penguins? How many columns?
```{r}
nrow(penguins)
ncol(penguins) 
```

- Answer:
  * 344 rows and 8 columns

2) What does the `bill_depth_mm` variable in the penguins dataframe describe?
```{r}
?penguins
```

- Answer:
  * `bill_depth_mm` is a number denoting bill depth in millimeters

3) Make a scatterplot of `bill_depth_mm` vs. `bill_length_mm`. That is, make a scatterplot with `bill_depth_mm` on the y-axis and `bill_length_mm` on the x-axis. Describe the relationship between these two variables.
```{r}
ggplot(
  penguins,
  aes(x = bill_length_mm, y = bill_depth_mm)
) +
  geom_point()
```

- Answer:
  * There seems to be no strong linear correlation between those variables
  * The data points apparently form 3 clusters
  * It seems that penguins with less deep bills tend to have longer bills

4) What happens if you make a scatter plot of species vs. bill_depth_mm? What might be a better choice of geom?
```{r}
ggplot(
  penguins,
  aes(x = species, y = bill_depth_mm)
) +
  geom_bar(stat = "identity")
```

- Answer:
  * Since `species` is a categorical variable, it goes in the X-axis
  * If we use `geom_point`, the scatter plot will have 3 vertical lines for each species
    * In this case, it is better to use a bar plot
    * However, we have to notice something when using the bar plot:
        * By default, `geom_bar` counts the number of rows for each x value
        * But, we want to count bill depth, so we have to use the `stat = "identity"`


5) What does the `na.rm` argument do in `geom_point()`? What is the default value of the argument? Create a scatterplot where you successfully use this argument set to `TRUE`.

```{r}
# Using "na.rm = TRUE"
ggplot(data = penguins,
       aes(x = flipper_length_mm, y = body_mass_g)) + 
  geom_point(na.rm = TRUE)
```

- Answer:
  * `na.rm` gets rid of missing values
  * The default is `na.rm = FALSE`, which is to say that R accounts for those `NAs`

6) Add the following caption to the plot you made in the previous exercise: “Data come from the palmerpenguins package".

- Answer:
```{r}
ggplot(data = penguins,
       aes(x = flipper_length_mm, y = body_mass_g)) + 
  geom_point(na.rm = TRUE) +
  labs(title = "Data come from the palmerpenguins package")
```

7) Recreate the following visualization. What aesthetic should `bill_depth_mm` be mapped to? And should it be mapped at the global level or at the geom level?

```{r}
ggplot(data = penguins,
       aes(x = flipper_length_mm, y = body_mass_g)) + 
  geom_point(aes(color = bill_depth_mm)) +
  geom_smooth() +
  labs(
    title = "Data come from the palmerpenguins package")
```

- Answer:
  * We should color by `bill_depth_mm` in the `geom_point() function`
     * When we map the color of the values of a continuous variable (numeric), we get a gradient of colors
     * It is different when we do it with a categorical variable (factor), where each category gets a distinct color
  * `bill_depth_mm` should not be mapped at the global level, since we want to fit a single line for the entire dataset
  * We do not need to specify the method in the geom_smooth, since we do not want a line


8) Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions.

- Answer:
```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g, color = island)
) +
  geom_point() +
  geom_smooth(se = FALSE)
```

9) Will these two graphs look different? Why/why not?
```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point() +
  geom_smooth()

ggplot() +
  geom_point(
    data = penguins,
    mapping = aes(x = flipper_length_mm, y = body_mass_g)
  ) +
  geom_smooth(
    data = penguins,
    mapping = aes(x = flipper_length_mm, y = body_mass_g)
  )
```

- Answer:
  *  They result in the same graph


# 4. Visualizing distributions

## A. Categorical Variable

- A variable is categorical if it can only take one of a small set of values
-  We can use a bar chart to see its distribution
  * The height of the bars displays the number of observations occurred with each X value
  
- Example:
```{r}
ggplot(penguins, aes(x = species)) +
  geom_bar()
```


## B. Numerical Variable:

- A variable is numerical if it can take on a wide range of numerical values
  * Numerical variables can be CONTINUOUS or DISCRETE
- The histogram is commonly used to visualize the distribution of CONTINUOUS variables:
```{r}
ggplot(penguins,
       aes(x = body_mass_g)) +
  geom_histogram(binwidth = 200)
```

- It divides the x-axis into equally spaced bins
  * The height of the bar displays the number of observations that fall in each bin
- Note that it is possible to alter the `binwidth`
  * This is good because different binwidths can reveal different patterns

- The Density Plot is an alternative way to visualize the distribution of numerical variables
```{r}
ggplot(penguins,
       aes(x = body_mass_g)) +
  geom_density()
```
- It is a smoothed-out version of a histogram
  * It is particularly good for continuous variables that come from a underlying smooth distribution


# 5. Exercises (2)

1) Make a bar plot of species of penguins, where you assign `species` to the y aesthetic. How is this plot different?
```{r}
ggplot(penguins,
       aes(y = species)) +
  geom_bar()
```

- Answer:
  * The graph is inverted -> the categories are in the Y axis and the count in the X axis.

2) How are the following 2 plots different? Which aesthetic, color or fill, is more useful for changing the color of bars?
```{r}
ggplot(penguins,
       aes(x = species)) +
  geom_bar(color = "red")

ggplot(penguins,
       aes(x = species)) +
  geom_bar(fill = "red")
```

- Answer:
  * The 2nd one is more useful, because it fills the whole bar with the color

3) Make a histogram of the `carat` variable in the `diamonds` dataset that is available when you load the `tidyverse` package. Experiment with different binwidths. What binwidth reveals the most interesting patterns?
```{r}
ggplot(diamonds,
       aes(x = carat)) +
  geom_histogram(binwidth = 0.2)
```

- Answer:
  * Since we have small-value observations in the X-axis, smaller binwidths are better
  
# 6. Visualizing relationships

## 6.1 Two numerical variables
- To compare 2 numerical variables, we use the scatter plot

## 6.2 Numerical and categorical variable
- Let's see the distribution of body mass (numerical) by species (categorical)
```{r}
ggplot(penguins,
       aes(x = species, y = body_mass_g)) +
  geom_boxplot()
```

## 6.3 Two categorical variables

- We can use stacked bar plots to see the relationship between 2 categorical variables
```{r}
ggplot(penguins,
       aes(x = island, fill = species)) +
  geom_bar()
```

- We see that there is the same amount of Adelies on each island
  * However, we do not have a good sense of the percentage
- To better compare species distributions across islands, we use a relative frequency plot

```{r}
ggplot(penguins,
       aes(x = island,  fill = species)) +
  geom_bar(position = "fill")

getwd()
```

- The `position = fill` argument changes the way the bar plot is displayed
  *  It stacks the bars for each group and then scales them so that the total height of each stacked bar is 1 (or 100%)
  * Each bar is showing the proportion of each subgroup within the main group
- In sum, to create these bar charts, map the variable for bar separation to the X aesthetic and the variable for bar colors to the `fill` aesthetic


## 6.4 Three or more variables

- At the some point, adding too many aesthetic mappings to a plot makes it cluttered and difficult to make sense of

- Another way, particularly useful for categorical variables, is to split your plot into facets
  * Facets are subplots that each display one subset of the data
  
- To facet your plot by a single variable, use `facet_wrap()`
```{r}
ggplot(penguins,
       aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species, shape = species)) +
  facet_wrap(~island)
```

- The first argument of `facet_wrap()` is a formula, which `you`can be created with `~`, followed by a variable name
  * The variable that you pass to `facet_wrap()` should be categorical


# 7. Exercises (3)

1) The `mpg` dataframe that is bundled with the `ggplot2` package contains 234 observations collected by the US Environmental Protection Agency on 38 car models. Which variables in `mpg` are categorical? Which variables are numerical? How can you see this information when you run `mpg`?
```{r}
glimpse(mpg)
```

- Answer:
  * The categorical variables are: `manufacturer`, `model`, `trans`, `fl`, `class`, `drv`
  * The numerical variables are: `dipl`, `year`, `cyl`, `cty`, `hwy`
  
2) Make a scatter plot of `hwy` vs. `displ` using the mpg dataframe. Next, map a third numerical variable to `color`, then `size`, then both `color` and `size`, then `shape`. How do these aesthetics behave differently for categorical vs. numerical variables?

- Answer:
  * Basic plot
```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point()
```

  * Color (numerical)
    * Continuous gradient
    * Used to show intensity or scale
```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = cty)) +
  geom_point()
```

  * Size (numerical)
    * Point size varies by value
    * Easy to compare relative magnitudes
```{r}
ggplot(mpg, aes(x = displ, y = hwy, size = cty)) +
  geom_point()
```

  * Color + Size (numerical)
    * Combines the color gradient and the difference sizes
    * Problem: it can be redundant
```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = cty, size = cty)) +
  geom_point()
```
  
  * The `shape` function work only with categorical variables

3) In the scatter plot of `hwy` vs. `displ`, what happens if you map a third variable to `linewidth`?
```{r}
ggplot(mpg,
       aes(x = displ, y = hwy, linewidth = cyl)) +
  geom_point()
```

- Answer:
  * Mapping a third variable to `linewidth` will have no effect, because it is not a valid aesthetic for points

4) What happens if you map the same variable to multiple aesthetics?
```{r}
ggplot(mpg,
       aes(x = displ, y = hwy, size = cyl, color = cyl)) +
  geom_point()
```

- Answer:
  * It can enhance clarity by reinforcing the variable’s role in the plot
  * However, it can also overcomplicate the plot

5) Make a scatter plot of `bill_depth_mm` vs. `bill_length_mm` and color the points by `species`. What does adding coloring by species reveal about the relationship between these two variables? What about faceting by species?
```{r}
ggplot(penguins,
       aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point(aes(color = species)) +
  facet_wrap(~species)
```

- Answer:
  * The relationship between bill depth and bill length varies according to species
    *  Gentoos and Chintraps have similar bill length, but Chinstrap have higher bill depth
    * Adelies have less bill length, but high bill depth

6) Why does the following yield two separate legends? How would you fix it to combine the two legends?
```{r}
ggplot(
  data = penguins,
  mapping = aes(
    x = bill_length_mm, y = bill_depth_mm, 
    color = species, shape = species
  )
) +
  geom_point() +
  labs(color = "Species")
```

- Answer:
  * When you map both `color` and `shape` to the same variable (e.g., `species`), `ggplot2` will combine their legends into a unified one
    * But, it will do it if and only if both legends have the same title (name)
```{r}
ggplot(
  data = penguins,
  mapping = aes(
    x = bill_length_mm, y = bill_depth_mm, 
    color = species, shape = species
  )
) +
  geom_point() +
  labs(color = "Species", shape = "Species")
```


7 Create the two following stacked bar plots. Which question can you answer with the first one? Which question can you answer with the second one?

```{r}
ggplot(penguins,
       aes(x = island, fill = species)) +
  geom_bar(position = "fill")

ggplot(penguins,
       aes(x = species, fill = island)) +
  geom_bar(position = "fill")
```

- Answer:
  * The 1st plot show the distribution of each species on each island
  * The 2nd plot show the distribution of each island on each species of penguin
  
# 8. Saving Plots

- We can use `ggsave` to save the plots in the directory
```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point()
ggsave(filename = "penguin-plot.png")
```

- The `ggsave` function will save the plot that comes right after it in your code
