---
title: "2. Workflow Basics"
author: "Matheus C. Negrellos"
date: "2025-07-21"
output: html_document
---

# 1. "Rules" for Comments

- A reasonable rule of thumb to follow when writing comments after the "#" is:
  *  Use comments to explain the WHY of your code, not the how or the what
- We can discover the how and the what by carefully reading the code
  * However, the WHY you've changed the value of an argument from its default is almost impossible to discover.
    * In this case, comments are welcomed.


# 2. Calling functions

```{r}
# "seq" stands for "sequence" of numbers
seq(from = 1, to = 10)

# Usually, the "from" and "to" are omitted:
seq(1, 10)
```

# 3. Exercises

1. Why does this code not work?
```{r}
my_variable <- 10
my_varıable
```

- Answer:
  * The second line has a typo error: the "i" is actually a dotless "i" (ı)


2. Tweak each of the following R commands so that they run correctly:
```{r}
# Original
libary(todyverse)

ggplot(dTA = mpg) + 
  geom_point(maping = aes(x = displ y = hwy)) +
  geom_smooth(method = "lm)"

# Corrected version:
library(tidyverse)

ggplot(mpg,
       aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth(method = "lm")
```

- Answer:
  * Besides the typo errors, there is a problem with the aesthetics mapping
    * If you do not map the `aes` at the global level (inside "ggplot"), the following `geoms` will not receive the x and y aesthetics
    * If you map the asesthetics only inside `geom_point`, `geom_smooth` does not know which variables to use, unless you specify them.


3. Press Option + Shift + K / Alt + Shift + K. What happens? How can you get to the same place using the menus?

- Answer:
  * This command opens a quick list of keyboard shortcuts that can also be opened by clicking in the "help" button

4. Let’s revisit an exercise from the Section 1.6. Run the following lines of code. Which of the two plots is saved as mpg-plot.png? Why?
```{r}
my_bar_plot <- ggplot(mpg, aes(x = class)) +
  geom_bar()

my_scatter_plot <- ggplot(mpg, aes(x = cty, y = hwy)) +
  geom_point()

ggsave(filename = "mpg-plot.png", plot = my_bar_plot)
```

- Answer:
  * The first plot is saved as "mpg-plot.png" because it is the one in the argument that indicates which object to save: `plot = my_bar_plot`
